---
title: "Week 7: Strings"
subtitle: "EMSE 6574, Section 11 "
author: "John Helveston"
date: October 07, 2019
output:
  xaringan::moon_reader:
    css: classNotes.css
---

# Announcments

If you haven't already, install the **stringr** library:
```{r eval=FALSE}
install.packages("stringr")
library(stringr)
```
```{r echo=FALSE}
library(stringr)
```

---
# Making a string

Single or double quotes - they both work:

```{r eval=FALSE}
s1 <- "This is a string"
s1
s2 <- 'I can use single quotes to include a "quote" inside a string'
s2
```

--
Use them where it makes sense, e.g.:
```{r, eval=FALSE}
cat("It's a boy!")
cat('I said, "Hi!"')
```

---
# String literals and "escaping"

Use the `\` symbol to "escape" a literal symbol:

--
```{r}
cat('\"') # Double quote
```
--
```{r}
cat('\'') # Single quote
```
--
```{r}
cat('This\\that') # Backslash
```
--
```{r}
cat('This\nthat') # New line
```
--
```{r}
cat('This\tthat') #Tab space
```

---
# String constants

R has a few built-in string constants:
--
```{r}
LETTERS
letters
month.abb
month.name
```

--
The **stringr** library has some longer string constants:
```{r}
head(words)
head(sentences)
head(fruit)
```

---
# Main **stringr** functions

Most start with "`str_`":

Function        | Description
----------------|-----------------------------------------
`str_length()`  | number of characters
`str_sub()`     | extracts substrings
`str_dup()`     | duplicates characters
`str_c()`       | string concatenation
`str_trim()`    | removes leading and trailing whitespace
`str_pad()`     | pads a string
`str_()`        | pads a string
`str_sort()`    | sort a string alphabetically
`str_order()`   | get the order of a sorted string
`str_split()`   | split a string into a vector
`str_detect()`  | match a string in another string
`str_replace()` | replace a string in another string

---
# `str_length()`: Get the number of characters in a string

```{r}
str_length(" ")
str_length("hello world")
str_length("The quick brown fox jumped over the lazy dog")
```

---
# `str_sub()`: Access characters by their index

```{r}
x <- "Apple"
str_sub(x, 1, 3)
# Negative numbers count backwards from the end
str_sub(x, -3, -1)
```

--
If string is too short, it won't error:
```{r}
str_sub("App", 1, 5)
```

--
Use `str_sub()` to modify a string:
```{r}
x <- 'abcdef'
str_sub(x, 1, 3) <- 'ABC'
x
```

---
# `str_dup()`: Like `rep()` but for strings

--
```{r}
str_dup("hola", 3)
```

--
Note the difference with `rep()` (which creates a vector):
```{r}
rep("hola", 3)
```

---
# Most `stringr` functions work on string vectors

```{r}
x <- c("apples", "oranges")
x
```

--
Get the 3rd letter in each string in the vector
```{r}
str_sub(x, 3, 3)
```

--
Duplicate each string twice
```{r}
str_dup(x, 2)
```

---
# `str_c()`: Combine strings into one string

--
```{r}
str_c('x', 'y', 'z')
```

--
Use the `sep` argument to control how they're separated:
```{r}
str_c('x', 'y', 'z', sep = "-")
```

--
Note the difference with _vectors_ of strings:
```{r}
x <- c('x', 'y', 'z')
str_c(x)
```
--
To make a single string from a vector of strings, use `collapse`:
```{r}
str_c(x, collapse = '')
```

---
# Practice: Think-Pair-Share

[60 seconds](https://www.youtube.com/watch?v=CH50zuS8DD0) - no typing!

```{r}
printGreeting <- function(name, timeOfDay, isBirthday) {
    greeting <- str_c(
        "Good ", timeOfDay, " ", name,
            if (isBirthday) {
                ", and HAPPY BIRTHDAY!"
            } else {
                '.'
            }
        )
    cat(greeting)
}
```
What will this print?
```{r, eval=FALSE}
printGreeting('John', 'morning', isBirthday = FALSE)
printGreeting('John', 'morning', isBirthday = TRUE)
```
```{r, echo=FALSE}
printGreeting('John', 'morning', isBirthday = FALSE)
printGreeting('John', 'morning', isBirthday = TRUE)
```

---
# `str_trim()`: Remove excess white space

```{r}
x <- c("  a   ", "b   ",  "   c")
x
```
--
```{r}
str_trim(x) # Default trims both sides
```
--
```{r}
str_trim(x, side = "left") # Only trim left side
```
--
```{r}
str_trim(x, side = "right") # Only trim right side
```

---
# `str_pad()`: Add white space

--
```{r}
x <- c("abc", "defghi")
x
```
--
```{r}
str_pad(x, 10) # Default pads on left
```
--
```{r}
str_pad(x, 10, side = "both") # Pad both sides
```
--
Pad with a different character:
```{r}
str_pad(x, 10, side = "both", pad = '*')
```

---
# Case conversion: `str_to_lower()`, `str_to_upper()`, `str_to_title()`

We saw these in HW 1!
```{r}
x <- "Want to hear a joke about paper? Never mind, it's tearable."
x
```
```{r}
str_to_lower(x)
str_to_upper(x)
str_to_title(x)
```

---
# Sort string vectors alphabetically

```{r}
x <- c('Y', 'M', 'C', 'A')
x
```
--
```{r}
str_sort(x)
```
--
```{r}
str_sort(x, decreasing = TRUE)
```
--
```{r}
str_order(x)
```
--
```{r}
x[str_order(x)]
```

---
# `str_split()`: Split a string into multiple strings

--
```{r}
x <- 'This string has spaces-and-dashes'
x
```
--
```{r}
str_split(x, " ") # Split on the spaces
```
--
```{r}
str_split(x, "-") # Split on the dashes
```

---
# What's with the `[[1]]` thing?

--
`str_split()` returns a `list` of vectors
--
Example:
```{r}
x <- c('babble', 'scrabble')
str_split(x, 'bb')
```
--
If you're only splitting one string, add `[[1]]` to get the first vector:
```{r}
str_split('hooray', 'oo')[[1]]
```

---
# Very common splits (you should memorize these)

--
Splitting on `""` breaks a string into _characters_:
```{r}
str_split("apples", "")[[1]]
```
--
Splitting on `" "` breaks a _sentence_ into words:
Break a sentence into individual words:
```{r}
x <- "If you want to view paradise, simply look around and view it"
str_split(x, " ")[[1]]
```

---
# `str_detect(string, pattern)`: Is `pattern` in `string`?

--
```{r}
tenFruit <- fruit[1:10]
tenFruit
```
--
```{r}
str_detect(tenFruit, "a")
```

--
How many in whole `fruit` vector contain the letters `"berry"`?:
```{r}
length(fruit)
sum(str_detect(fruit, "berry")) # Raw count
sum(str_detect(fruit, "berry")) / length(fruit) # Percentage of total
```

---
# Detect if string starts / ends with pattern

--
- Insert `^` to match _start_ of string.
- Insert `$` to match _end_ of string.

--
Example:
```{r}
tenFruit
# Which fruit start with "a"?
str_detect(tenFruit, "^a")
# Which fruit end with "y"?
str_detect(tenFruit, "y$")
```

--
Trick to remember this:

### If you _start_ with power (`^`), you'll _end_ up with money (`$`).

---
# `str_count()`: Count number of times pattern appears

--
Example:
```{r}
x <- c("apple", "banana", "pear")
str_count(x, "a")
```

--
Note the difference with `str_detect()`:
```{r}
str_detect(x, "a")
```
---
# `str_replace()`: Replace matched strings with new string

--
Example:
```{r}
x <- c("apple", "pear", "banana")
```
--
```{r}
str_replace(x, "a", "-")
```
--
```{r}
str_replace_all(x, "a", "-")
```











---

# Practice

`getUniqueChars(s)`

Write a function that takes a single string, `s`, and returns a _sorted_ vector of the unique characters the string below. All letters should be lowercase (so `"A"` should be treated the same as `"a"`).

Example:
`s <- 'babbleScrabbleApple'`
`getUniqueChars(s) == c("a","b","c","e","l","p","r","s")`

Hints:

- Use `str_split()` to break a string into characters
- Check out the `unique()` function

--
```{r}
s <- 'babbleScrabbleApple'
s <- str_to_lower(s)
chars <- str_split(s, "")[[1]] # Split the string into characters
uniqueChars <- unique(chars)
str_sort(uniqueChars)
```















---
# Group practice

[20 minutes](https://www.youtube.com/watch?v=kxGWsHYITAw) - In groups of 4, write the following functions:

.pull-left[
1) `reverseString(s)`

Write a function that returns the string in reverse order. So if `s` equals `"abcde"`,  `reverseString(s)` should equal `"edcba"`. You may assume that `s` only contains upper and/or lower case letters, but your solution must correctly return capital letters in their appropriate order. Here's some test cases:

- `reverseString("aWordWithCaps") == "spaChtiWdroWa"`
- `reverseString("abcde") == "edcba"`
- `reverseString("") == ""`
- `reverseString("aa") == "aa"`
]

.pull-right[
2) `isPalindrome(s)`

Write a function that returns `TRUE` if the string `s` is a [Palindrome](https://en.wikipedia.org/wiki/Palindrome) and `FALSE` otherwise. The string `s` can contains any letter, number, or symbol, but it will be a character data type. Here's some test cases:

- `isPalindrome("abcba") == TRUE`
- `isPalindrome("abcb") == FALSE`
- `isPalindrome("321123") == TRUE`

]

---
# Group practice

[20 minutes](https://www.youtube.com/watch?v=kxGWsHYITAw) - In groups of 4, write the following functions:

.pull-left[
1) `sortString(s)`

Write the function `sortString(s)` that takes a string `s` and returns back an alphabetically sorted string. So `sortString("cba")` should return `"abc"`. You may assume that `s` only contains upper and/or lower case letters. Here's some test cases:

- `sortString("cba") == "abc"`
- `sortString("abedhg") == "abdegh"`
- `sortString("AbacBc") == "aAbBcc"`

]

.pull-right[
2) `areAnagrams(s1, s2)`

**Background**: An [anagram](https://en.wikipedia.org/wiki/Anagram) is a word or phrase formed by rearranging the letters of a different word or phrase, typically using all the original letters exactly once. With that in mind, write the function `areAnagrams(s1, s2)` that takes two strings, `s1` and `s2` (that you may assume contain only upper and/or lower case letters), and returns `TRUE` if the strings are anagrams, and `FALSE` otherwise. Treat `"a"` and `"A"` as the same letters (so `"Aba"` and `"BAA"` are anagrams). Hint: you can use either `sortString(s)` or `letterCount(s)` as a helper function - each are helpful for different approaches to solving this. Here's some test cases:

- `areAnagrams("", "") == TRUE`
- `areAnagrams("abCdabCd", "abcdabcd") == TRUE`
- `areAnagrams("abcdaBcD", "AAbbcddc") == TRUE`
- `areAnagrams("abcdaabcd", "aabbcddcb") == FALSE`

]



