---
title: "Week 7: Strings"
subtitle: "EMSE 6574, Section 11 "
author: "John Helveston"
date: October 07, 2019
output:
  xaringan::moon_reader:
    css: classNotes.css
---

# Quiz 3

### [20 minutes](https://www.youtube.com/watch?v=kxGWsHYITAw)

- ### No calculators
- ### No notes
- ### No books
- ### No computers
- ### No phones

---
# Announcements

--
1) If you haven't already, install the **stringr** library:
```{r eval=FALSE}
install.packages("stringr")
library(stringr)
```
```{r echo=FALSE}
library(stringr)
```
<br>

--
2) You have **3** weeks for HW 4
<br>
<br>

--
3) **This week**: Instead of office hours, we'll have an exam review

---
# Making a string

Single or double quotes - they both work:

```{r}
cat("This is a string")
cat('This is a string')
```
<br>

--
Use them where it makes sense, e.g.:
```{r}
cat("It's a boy!")
cat('I said, "Hi!"')
```

---
# Making a string

What if a string has both `'` and `"` symbols?
<br>

Example: `It's nice to say, "Hi!"`

--
```{r, error = TRUE}
cat("It's nice to say, "Hi!"")
```
--
```{r, error = TRUE}
cat('It's nice to say, "Hi!"'')
```

---
# "Escaping" to the rescue!

--
Use the `\` symbol to "escape" a literal symbol:
--
```{r}
cat("It's nice to say, \"Hi!\"") # Double quote
```
--
```{r}
cat('It\'s nice to say, "Hi!"') # Single quote
```
<br>
--
```{r}
cat('This\nthat') # New line
```
--
```{r}
cat('This\tthat') # Tab space
```
--
```{r}
cat('This\\that') # Backslash
```

---
# String constants

R has a few built-in string constants:
--
```{r}
LETTERS
```
--
```{r}
letters
```
--
```{r}
month.name
```
--
```{r}
month.abb
```

---
# String constants

The **stringr** library has some longer string constants:
--
```{r}
head(words)
length(words)
```
--
```{r}
head(sentences)
```
--
```{r}
head(fruit)
```

---
# Main **stringr** functions

|Function          |  Description                            |
|:-----------------|:----------------------------------------|
|`str_to_lower(x)` | converts string to lower case           |
|`str_to_upper(x)` | converts string to upper case           |
|`str_to_title(x)` | converts string to title case           |
|`str_length()`    | number of characters                    |
|`str_sub()`       | extracts substrings                     |
|`str_dup()`       | duplicates characters                   |
|`str_trim()`      | removes leading and trailing whitespace |
|`str_pad()`       | pads a string                           |
|`str_c()`         | string concatenation                    |
|`str_split()`     | split a string into a vector            |
|`str_sort()`      | sort a string alphabetically            |
|`str_order()`     | get the order of a sorted string        |
|`str_detect()`    | match a string in another string        |
|`str_replace()`   | replace a string in another string      |

---
# Case conversion

We saw these in HW 1!
```{r}
x <- "Want to hear a joke about paper? Never mind, it's tearable."
```
--
```{r}
str_to_lower(x)
```
--
```{r}
str_to_upper(x)
```
--
```{r}
str_to_title(x)
```
--
```{r}
library(tools)
toTitleCase(x)
```

---
# Comparing strings

```{r}
a <- "Apples"
b <- "apples"
a == b
```
<br>

--
Use `str_to_lower()` or `str_to_upper()` to **ignore case**:
<br>
<br>

```{r}
str_to_lower(a) == str_to_lower(b)
str_to_upper(a) == str_to_upper(b)
```

---
# Get the number of characters in a string

`str_length()`
<br>
<br>

--
```{r}
str_length("hello world")
```
--
```{r}
str_length("The quick brown fox jumped over the lazy dog")
```
--
```{r}
str_length(" ")
```
--
```{r}
str_length("")
```

---
# Access characters by their index

`str_sub()`
<br>
<br>

--
```{r}
str_sub("Apple", 1, 3)
```

--
```{r}
str_sub("Apple", -3, -1) # Negative numbers count backwards from the end
```

--
```{r}
str_sub("hi", 1, 5) # If string is too short, R won't error
```
<br>

--
Use `str_sub()` to modify a string:
```{r}
x <- 'abcdef'
str_sub(x, 1, 3) <- 'ABC'
```
--
```{r}
x
```

---
# Repeat a string

`str_dup()`
<br>
<br>

--
```{r}
str_dup("hola", 3)
```
<br>

--
Note the difference with `rep()`:
```{r}
rep("hola", 3)
```

---
# `stringr` functions work on vectors

--
```{r}
x <- c("apples", "oranges")
x
```
<br>

--
Get the first 3 letters in each string:
```{r}
str_sub(x, 1, 3)
```
<br>

--
Duplicate each string twice
```{r}
str_dup(x, 2)
```
<br>

--
```{r}
str_to_upper(x)
```

---
# Quick practice: Think-Pair-Share

1) Create the string `x <- 'thisIsGoodPractice'`

2) Use `stringr` functions to transform `x` into the following strings:

- `'thisIsGood'`
- `'practice'`
- `'GOOD'`
- `'thisIsGoodPracticethisIsGoodPractice'`
- `'thisthisthis'`
- `'GOODGOODGOOD'`

**Hint**: You'll need these:
- `str_to_lower()`
- `str_to_upper()`
- `str_sub()`
- `str_dup()`

---
# Remove excess white space

`str_trim()`
<br>
<br>

--
```{r}
x <- "         aStringWithSpace        "
x
```
<br>

--
```{r}
str_trim(x) # Trims both sides by default
```
--
```{r}
str_trim(x, side = "left") # Only trim left side
```
--
```{r}
str_trim(x, side = "right") # Only trim right side
```

---
# Add white space (or other characters)

`str_pad()`
<br>
<br>

--
```{r}
x <- "hello"
x
```
<br>

--
```{r}
str_pad(x, width = 10) # Inserts pad on left by default
```
--
```{r}
str_pad(x, width = 10, side = "both") # Pad both sides
```
--
Pad with a different character:
```{r}
str_pad(x,  width = 10, side = "both", pad = '*')
```

---
# Combine strings into one string

`str_c()`
<br>
<br>

--
```{r}
str_c('x', 'y', 'z')
```

--
Control separation with `sep` argument:
```{r}
str_c('x', 'y', 'z', sep = "-")
```

--
Note the difference with _vectors_ of strings:
```{r}
x <- c('x', 'y', 'z')
str_c(x)
```
--
To make a single string from a vector of strings, use `collapse`:
```{r}
str_c(x, collapse = "")
```

---
# `str_c` works with function logic

```{r}
printGreeting <- function(name, timeOfDay, isBirthday) {
    greeting <- str_c(
        "Good ", timeOfDay, " ", name,
            if (isBirthday) {
                ", and HAPPY BIRTHDAY!"
            } else {
                '.'
            }
        )
    cat(greeting)
}
```
What do you think this will print?
```{r, eval=FALSE}
printGreeting('John', 'morning', isBirthday = FALSE)
printGreeting('John', 'morning', isBirthday = TRUE)
```
--
```{r, echo=FALSE}
printGreeting('John', 'morning', isBirthday = FALSE)
printGreeting('John', 'morning', isBirthday = TRUE)
```

---
# Split a string into multiple strings

`str_split()`
<br>
<br>

--
```{r}
x <- 'This string has spaces-and-dashes'
x
```
<br>

--
```{r}
str_split(x, " ") # Split on the spaces
```
--
```{r}
str_split(x, "-") # Split on the dashes
```

---
# What's with the `[[1]]` thing?

--
`str_split()` returns a `list` of vectors

--
Example:
```{r}
x <- c('babble', 'scrabblebabble')
str_split(x, 'bb')
```
<br>

--
If you're only splitting one string, add `[[1]]` to get the first vector:
```{r}
str_split('hooray', 'oo')[[1]]
```

---
# Common splits (memorize these)

--
Splitting on `""` breaks a string into _characters_:
```{r}
str_split("apples", "")[[1]]
```
<br>

--
Splitting on `" "` breaks a _sentence_ into words:
```{r}
x <- "If you want to view paradise, simply look around and view it"
str_split(x, " ")[[1]]
```

---
# Quick practice: Think-Pair-Share

1) Create the following objects:

```{r, eval=FALSE}
x <- 'this_is_good_practice'
y <- c('hello', 'world')
```

2) Use `stringr` functions to transform `x` and `y` into the following:

- `"hello world"`
- `"***hello world***"`
- `c("this", "is", "good", "practice")`
- `"this is good practice"`
- `"hello world, this is good practice"`

**Hint**: Create intermediate objects! And use these:
- `str_trim()`
- `str_pad()`
- `str_c()`
- `str_split()`

---
# Sort string vectors alphabetically

`str_sort()`
<br>
<br>

```{r}
x <- c('Y', 'M', 'C', 'A')
x
```
--
```{r}
str_sort(x)
```
--
```{r}
str_sort(x, decreasing = TRUE)
```
--
```{r}
str_order(x)
```
--
```{r}
x[str_order(x)]
```

---
# Detect if pattern is in string

`str_detect(string, pattern)`
<br>
<br>

--
```{r}
tenFruit <- fruit[1:10]
tenFruit
```
<br>

--
```{r}
str_detect(tenFruit, "berry")
```
<br>

--
Count how many have the string `"berry"`:
```{r}
sum(str_detect(tenFruit, "berry"))
```

---
# Count number of times pattern appears

`str_count(string, pattern)`
<br>
<br>

--
Example:
```{r}
x <- c("apple", "banana", "pear")
str_count(x, "a")
```
<br>

--
Note the difference with `str_detect()`:
```{r}
str_detect(x, "a")
```

---
# Detect if string _starts_ with pattern

Example: Which fruit _start_ with "a"?

--
```{r}
tenFruit <- fruit[1:10]
tenFruit
```
<br>

--
```{r}
str_detect(tenFruit, "a")
```
<br>

--
```{r}
str_detect(tenFruit, "^a")
```

---
# Detect if string _ends_ with pattern

Example: Which fruit _end_ with an "e"?

--
```{r}
tenFruit <- fruit[1:10]
tenFruit
```
<br>

--
```{r}
str_detect(tenFruit, "e")
```
<br>

--
```{r}
str_detect(tenFruit, "e$")
```

---
# Trick to remember this

### If you _start_ with power (`^`), you'll _end_ up with money (`$`).
<br>
<br>

--
```{r}
tenFruit
```
<br>

```{r}
str_detect(tenFruit, "^a")
str_detect(tenFruit, "e$")
```

---
# Quick practice: Think-Pair-Share

1) For these questions, we'll use the `fruit` vector:

```{r, eval=FALSE}
head(fruit)
```

2) Use `stringr` functions to answer the following questions about the `fruit` vector:

- How many fruit have the string "berry" in it?
sum(str_detect(fruit, "berry"))
- Which fruit contain more than one `o` characters?
fruit[str_count(fruit, "o") > 1]

**Hint**: You'll need these:
- `str_detect()`
- `str_count()`







---
# `str_replace()`: Replace matched strings with new string

--
Example:
```{r}
x <- c("apple", "pear", "banana")
```
--
```{r}
str_replace(x, "a", "-")
```
--
```{r}
str_replace_all(x, "a", "-")
```


---
# Practice: Think-Pair-Share








---
# Practice: Think-Pair-Share

`getUniqueChars(s)`

Write a function that takes a single string, `s`, and returns a _sorted_ vector of the unique characters the string below. All letters should be lowercase (so `"A"` should be treated the same as `"a"`).

Example:
```{r, eval=FALSE}
s <- 'babbleScrabbleApple'
getUniqueChars(s) == c("a","b","c","e","l","p","r","s")
```

Hints:

- Use `str_split()` to break a string into characters
- Check out the `unique()` function

--
```{r, eval=FALSE}
getUniqueChars <- function(s) {
    s <- str_to_lower(s)
    chars <- str_split(s, "")[[1]] # Split the string into characters
    return(str_sort(unique(chars)))
}
```








---
# Group practice

[20 minutes](https://www.youtube.com/watch?v=kxGWsHYITAw) - In groups of 4, write the following functions:

.pull-left[
1) `reverseString(s)`

Write a function that returns the string in reverse order. So if `s` equals `"abcde"`,  `reverseString(s)` should equal `"edcba"`. You may assume that `s` only contains upper and/or lower case letters, but your solution must correctly return capital letters in their appropriate order. Here's some test cases:

- `reverseString("aWordWithCaps") == "spaChtiWdroWa"`
- `reverseString("abcde") == "edcba"`
- `reverseString("") == ""`
- `reverseString("aa") == "aa"`
]

.pull-right[
2) `isPalindrome(s)`

Write a function that returns `TRUE` if the string `s` is a [Palindrome](https://en.wikipedia.org/wiki/Palindrome) and `FALSE` otherwise. The string `s` can contains any letter, number, or symbol, but it will be a character data type. Here's some test cases:

- `isPalindrome("abcba") == TRUE`
- `isPalindrome("abcb") == FALSE`
- `isPalindrome("321123") == TRUE`

]

---
# Group practice

[20 minutes](https://www.youtube.com/watch?v=kxGWsHYITAw) - In groups of 4, write the following functions:

.pull-left[
1) `sortString(s)`

Write the function `sortString(s)` that takes a string `s` and returns back an alphabetically sorted string. So `sortString("cba")` should return `"abc"`. You may assume that `s` only contains upper and/or lower case letters. Here's some test cases:

- `sortString("cba") == "abc"`
- `sortString("abedhg") == "abdegh"`
- `sortString("AbacBc") == "aAbBcc"`

]

.pull-right[
2) `areAnagrams(s1, s2)`

**Background**: An [anagram](https://en.wikipedia.org/wiki/Anagram) is a word or phrase formed by rearranging the letters of a different word or phrase, typically using all the original letters exactly once. With that in mind, write the function `areAnagrams(s1, s2)` that takes two strings, `s1` and `s2` (that you may assume contain only upper and/or lower case letters), and returns `TRUE` if the strings are anagrams, and `FALSE` otherwise. Treat `"a"` and `"A"` as the same letters (so `"Aba"` and `"BAA"` are anagrams). Hint: you can use either `sortString(s)` or `letterCount(s)` as a helper function - each are helpful for different approaches to solving this. Here's some test cases:

- `areAnagrams("", "") == TRUE`
- `areAnagrams("abCdabCd", "abcdabcd") == TRUE`
- `areAnagrams("abcdaBcD", "AAbbcddc") == TRUE`
- `areAnagrams("abcdaabcd", "aabbcddcb") == FALSE`

]
