---
title: "Week 6: Vectors"
subtitle: "EMSE 6574, Section 11 "
author: "John Helveston"
date: September 30, 2019
output:
  xaringan::moon_reader:
    css: classNotes.css
---

# We've been dealing with vectors all along!

```{r}
x <- 1
x
```

--
```{r}
is.vector(x)
```

---
# The "concatenate" function

The universal vector generator: `c()`

--
```{r}
x <- c(1, 2, 3)
x
```
--
```{r}
y <- c('one', 'two', 'three')
y
```

---
# Other ways to make a vector

1) Sequences (we saw these last week):

--
```{r}
seq(1, 5)
```
--
```{r}
1:5
```
<br>

--
2) Repeating a value:

--
```{r}
rep(5, 10)
```
--
```{r}
rep("snarf", 5)
```

---
# Elements in vectors must be the same type

--
If a vector has a single character element, R makes everything a **character**:
```{r}
c(1, 2, "3")
c(TRUE, FALSE, "TRUE")
```

--
If a vector has numeric and logical elements, R makes everything a **number**:
```{r}
c(1, 2, TRUE, FALSE)
```

--
If a vector has integers and floats, R makes everything a **float**:
```{r}
c(1L, 2, pi)
```

---
# Practice: What will this return?

[60 seconds](https://www.youtube.com/watch?v=CH50zuS8DD0) - no typing!


```{r, eval=FALSE}
f <- function(x) {
    m = x
    l = x + 4
    m = m + 5
    return(c(m, l))
}

x <- c(1, 2, 3)
y <- c(3, 4, 5)
f(x)
f(y)
```

---
# Doing math on numeric vectors

```{r}
x <- 1:10
x
```

--
Sum all the elements:

Using a loop:
```{r}
# Using a loop:
total <- 0
for (i in x) {
    total <- total + x[i]
}
total
```

--
Using a vector:
```{r}
# Using a vector operation:
sum(x)
```

-- 
Other useful vector functions:
```{r}
prod(x) # Product of all the elements
mean(x)
median(x)
max(x)
min(x)
```

---
# Non-numeric vectors

--
Character vectors:

```{r}
stringVector <- c('oh', 'what', 'a', 'beautiful', 'morning')
stringVector
typeof(stringVector)
```

--
Logical vectors

```{r}
logicalVector <- c(rep(TRUE, 3), rep(FALSE, 3))
logicalVector
```

--
Sometimes looks can be deceiving:

```{r}
y <- c(TRUE, FALSE, 'a_string')
y
typeof(y)
```

---
# Vector slicing

1. Use brackets `[]` and indices:
```{r}
x <- seq(1, 10)
x[1] # Returns the first element
x[3] # Returns the third element
x[1:3]  # Returns the first three elements
x[c(2, 7)] # Returns the 2nd and 7th elements
```

2. Use negative integers to _remove_ elements
```{r}
x[-1] # Returns everything except the first element
x[-c(2, 7)] # Returns everything except the 2nd and 7th elements
```

---
# Characters indices

You can name vector elements by

--
1. Using the `names()` function:
```{r}
x <- seq(1, 10)
names(x) <- c('one', 'two', 'three', 'four', 'five', 'six', 'seven', 'eight', 'nine', 'ten')
x
```

--
2. Using the `c()` function:
```{r}
y <- c('one' = 1, 'two' = 2, 'three' = 3)
y
```

--
You can use names to slice vectors:
```{r}
x['one'] 
y[c('two', 'three')] 
```

---
# Logical indices

--
```{r}
x <- seq(10)
x
```

--
```{r}
x > 5 # Create logical vector
```

--
```{r}
x[x > 5] # Put logical vector in brackets to filter out the TRUE elements
```

--
Another way - use the `which()` function:
```{r}
which(x < 5) # Returns indices of TRUE elements
x[which(x < 5)] # Use which to select elements based on a condition
```

---
# Practice: What will this return?

[60 seconds](https://www.youtube.com/watch?v=CH50zuS8DD0) - no typing!

```{r, eval=FALSE}
f <- function(x) {
    for (i in seq(length(x))) {
        x[i] <- x[i] + sum(x) + max(x)
    }
    return(x)
}
x <- c(2, 1, 0)
y <- c(1, 4, 7)
f(x)
f(y)
```

---
# Comparing vectors

Check if 2 vectors are the same:

```{r}
x <- c(1, 2, 3)
y <- c(1, 2, 3)
x == y
```

--
To check if the _whole_ vectors are the same, use `all()`:
```{r}
all(x == y)
```

--
To check if any elements are the same, use `any()`:
```{r}
a <- c(1, 2, 3)
b <- c(-1, 2,-3)
a == b
any(a == b)
```

---
# `all()` vs. `identical()`

--
`all()` only compares the _elements_ in vectors - it's looking for `TRUE` and `FALSE` values

--
`identical()` compares vector elements and _attributes_, like names:

```{r}
names(x) <- c('a', 'b', 'c')
names(y) <- c('one', 'two', 'three')
all(x == y) # Only compares the elements
identical(x, y) # Also compares the **names** of the elements
```

---
# Vectorized operations

```{r}
x1 <- c(1, 2, 3)
x2 <- c(4, 5, 6)
```
```{r}
# Addition
x1 + x2 # Returns (1+4, 2+5, 3+6)

# Subtraction
x1 - x2 # Returns (1-4, 2-5, 3-6)

# Multiplicattion
x1 * x2 # Returns (1*4, 2*5, 3*6)

# Addition
x1 / x2 # Returns (1/4, 2/5, 3/6)
```

---
# Vectorized operations

### Be careful - dimensions much match!

```{r, error=TRUE}
# Mis-matched dimensions only give a warning!
x1 <- c(1, 2, 3)
x2 <- c(4, 5)
x1 + x2
```

--
Single-value vector operations
```{r}
# A :
x1 <- c(1, 2, 3)
x2 <- c(4)
x1 + x2
```

---
# Sorting

```{r}
a = c(2, 4, 6, 3, 1, 5)
sort(a)
sort(a, decreasing = TRUE)
```

--
Use `order()` to get the index values of the sorted order:
```{r}
order(a)
```

--
This does the same thing as `sort(a)`:
```{r}
a[order(a)] 
```

---
# Using vectors instead of a loop

Example: Get the remainder for each value in a vector:

--
Using a loop:
```{r}
x <- c(3.1415, 1.618, 2.718)
for (i in 1:length(x)) {
    remainder <- x[i] %% 1
    print(remainder)
}
```

--
Using a vector:
```{r}
x %% 1
```

---
# Group practice 

[20 minutes](https://www.youtube.com/watch?v=kxGWsHYITAw) - In groups of 4, write the following functions:

.pull-left[
1) `reverseVector(x)`

Write a function that returns the vector in reverse order. So if `x` equals `c(2,3,4)`, then `reverseVector(x)` should equal `c(4,3,2)`. You cannot use the `rev()` function (which does exactly this). **No loops!** Here's some test cases:

- `reverseVector(c(5, 1, 3)) == c(3, 1, 5)`
- `reverseVector(c(FALSE, TRUE, TRUE)) == c(TRUE, TRUE, FALSE)`
- `reverseVector(seq(10)) == seq(10, 1, -1)`

]

.pull-right[
2) `dotProduct(a, b)`

**Background**: the "dot product" of two vectors is the sum of the products of the corresponding terms. So the dot product of the vectors `c(1,2,3)` and `c(4,5,6)` is `(1*4) + (2*5) + (3*6)`, or `4 + 10 + 18 = 32`.  

With this in mind, write a function that takes two vectors and returns the dot product of those vectors. If the vectors are not equal length, ignore the extra elements in the longer vector. **No loops!** Here's some test cases:

- `dotProduct(c(1,2,3), c(4,5,6)) == 32`
- `dotProduct(c(1,2), c(4,5,6)) == 14`
- `dotProduct(c(2,3,4), c(-7,1,9)) == 25`
- `dotProduct(c(0,0,0), c(-7,1,9)) == 0`

]

---
# Group practice 

[20 minutes](https://www.youtube.com/watch?v=kxGWsHYITAw) - In groups of 4, write the following functions:

.pull-left[
1) `alternatingSum(a)`

Write a function that takes a vector of numbers `a` and returns the alternating sum, where the sign alternates from positive to negative or vice versa. **No loops!** Here's some test cases:

- `alternatingSum(c(5,3,8,4)) == (5 - 3 + 8 - 4)`
- `alternatingSum(c(1,2,3)) == (1 - 2 + 3)`
- `alternatingSum(c(0,0,0)) == 0`
- `alternatingSum(c(-7,5,3)) == (-7 - 5 + 3)`

]

.pull-right[
2) `median(a)`

Write a function that takes a vector of numbers `a` and returns the median value in the vector, which is the value of the middle element (or the average of the two middle elements). **No loops!** Here's some test cases:

- `median(c(5,3,8,4)) == mean(c(3,8))`
- `median(c(0,0,0)) == 0`
- `median(c(1,2,3)) == 2`
- `median(c(4,5,6,7,8)) == 6`
- `median(c(4,5,6,7)) == mean(c(5,6))`

]
