---
title: "Week 6: Vectors"
subtitle: "EMSE 6574, Section 11 "
author: "John Helveston"
date: September 30, 2019
output:
  xaringan::moon_reader:
    css: classNotes.css
---

# Announcements

--
### 2nd tutor session on Sat. this week (not Fri.)

--
### Quiz 3 next week, Exam 1 week after that

--
### Quiz 2 review

---
# Iteration review: loop over a sequence
```{r}
seq(5)
```

--
.pull-left[
### `for` loop
<br>

```{r}
for (i in seq(5)) {
    cat(i, '\n')
}
```
]
--
.pull-right[
### `while` loop
<br>

```{r}
i <- 1
while (i <= 5) {
    cat(i, '\n')
    i <- i + 1
}
```
]

---
# Search for something in a sequence

```{r}
seq(5)
```

### Count of **even** numbers in sequence:

--
.pull-left[
### `for` loop
<br>

```{r}
count <- 0
for (i in seq(5)) {
    if (i %% 2 == 0) {
        count <- count + 1
    }
}
count
```
]
--
.pull-right[
### `while` loop
<br>

```{r}
count <- 0
i <- 1
while (i <= 5) {
    if (i %% 2 == 0) {
        count <- count + 1
    }
    i <- i + 1
}
count
```
]

---
# This week: Vectors!

### Note: We've been dealing with vectors all along!

```{r}
x <- 1
x
```

--
```{r}
is.vector(x)
```

--
```{r}
length(x)
```

---
# The "concatenate" function

The universal vector generator: `c()`

--
```{r}
# Numeric vectors
x <- c(1, 2, 3)
x
```
--
```{r}
# Character vectors
y <- c('one', 'two', 'three')
y
```
--
```{r}
# Logical vectors
y <- c(TRUE, FALSE, TRUE)
y
```

---
# Other ways to make a vector

1) Sequences (we saw these last week):

--
```{r}
seq(1, 5)
```
--
```{r}
1:5
```
<br>

--
2) Repeating a value:

--
```{r}
rep(5, 10)
```
--
```{r}
rep("snarf", 5)
```

---
# Repeating a vector

--
```{r}
x <- rep(c("foo", "snarf"), 3)
x
length(x)
```
<br>

--
Note the difference when adding the `each` argument:
```{r}
x <- rep(c("foo", "snarf"), each = 3)
x
length(x)
```

---
# Elements in vectors must be the same type

--
If a vector has a _single_ character element, R makes everything a **character**:
--
```{r}
c(1, 2, "3")
```
--
```{r}
c(TRUE, FALSE, "TRUE")
```
<br>

--
If a vector has numeric and logical elements, R makes everything a **number**:
--
```{r}
c(1, 2, TRUE, FALSE)
```
<br>

--
If a vector has integers and floats, R makes everything a **float**:
```{r}
c(1L, 2, pi)
```

---
# Doing math on numeric vectors

```{r}
x <- 1:10
x
```

--
```{r}
x + 5
```

--
```{r}
x / 2
```

--
```{r}
x + x
```

--
```{r}
x - x
```

---
# Practice: Think-Pair-Share

[3 minutes](https://www.youtube.com/watch?v=iHdviZkM7S4) - no typing!

.pull-left[
```{r}
f1 <- function(x) {
    m = x
    n = x + 4
    m = m + 5
    return(c(m, n))
}
```
]
.pull-right[
```{r}
f2 <- function(x) {
    return(c(x, x / 2))
}
```
]
<br>

What will this return?
```{r, echo=FALSE}
x <- c(1, 2, 3)
y <- c(TRUE, FALSE, 7)
```
```{r, eval=FALSE}
x <- c(1, 2, 3)
y <- c(TRUE, FALSE, 7)
f1(x)
f2(y)
```
--
```{r}
f1(x)
f2(y)
```

---
# Using vectors instead of a loop: Summation

```{r}
x <- 1:10
x
```

--
**Using a loop**:
```{r}
total <- 0
for (item in x) {
    total <- total + item
}
total
```
<br>

--
**Using the `sum()` function**:
```{r}
sum(x)
```

---
# Other useful "summary" functions

--
```{r}
x
```
--
```{r}
length(x)
```
--
```{r}
prod(x)
```
--
```{r}
mean(x)
```
--
```{r}
median(x)
```
--
```{r}
max(x)
```
--
```{r}
min(x)
```

---
# Using vectors instead of a loop

Example: Get the remainder for each value in a vector:

```{r}
x <- c(3.1415, 1.618, 2.718)
x
```
--
**Using a loop**:
```{r}
remainder <- c()
for (i in x) {
    remainder <- c(remainder, i %% 1)
}
remainder
```

--
**Using a vector**:
```{r}
remainder <- x %% 1
remainder
```

---
# Get elements in a vector: Use brackets `[]`

```{r}
x <- seq(1, 10)
x
```
--
```{r}
x[1] # Returns the first element
```
--
```{r}
x[3] # Returns the third element
```
--
```{r}
x[1:3]  # Returns the first three elements
```
--
```{r}
x[c(2, 7)] # Returns the 2nd and 7th elements
```
--
```{r}
x[length(x)] # Returns the last element
```

---
# Use negative integers to _remove_ elements

```{r}
x <- seq(1, 10)
x
```
--
```{r}
x[-1] # Returns everything except the first element
```
--
```{r}
x[-1:-3]  # Returns everything except the first three elements
```
--
```{r}
x[-c(2, 7)] # Returns everything except the 2nd and 7th elements
```
--
```{r}
x[-length(x)] # Returns everything except the last elements
```

---
# You can name vector elements

```{r}
x <- seq(5)
x
```

--
1) Use the `names()` function:
```{r}
names(x) <- c('a', 'b', 'c', 'd', 'e')
x
```

--
2) Use the `c()` function:
```{r}
y <- c('a' = 1, 'b' = 2, 'c' = 3, 'd' = 4, 'e' = 5)
y
```

---
# You can use names to extract elements in a vector

```{r}
x
```
--
```{r}
x['a']
```
--
```{r}
x[c('a', 'c')]
```

---
# Logical indices

--
```{r}
x <- seq(10, 1)
x
```

--
Create a logical vector:
```{r}
x > 5
```

--
Use logical vector in brackets to filter elements:
```{r}
x[x > 5]
```

---
# Practice: Think-Pair-Share

[4 minutes](https://www.youtube.com/watch?v=zVHWhLme2NQ) - no typing!

```{r}
f <- function(x) {
    for (i in seq(length(x))) {
        x[i] <- x[i] + sum(x) + max(x)
    }
    return(x)
}
x <- c(1, 2, 3)
```

What will this return?
```{r, eval=FALSE}
f(x)
```
--
```{r, echo=FALSE}
f(x)
```

---
# Comparing vectors

Check if 2 vectors are the same:

```{r}
x <- c(1, 2, 3)
y <- c(1, 2, 3)
```
```{r, eval=FALSE}
x == y
```
--
```{r, echo=FALSE}
x == y
```
<br>

--
To check if _all_ elements are the same, use `all()`:
```{r}
all(x == y)
```

---
# Comparing vectors

To check if _any_ elements are the same, use `any()`:

--
```{r}
a <- c(1, 2, 3)
b <- c(-1, 2,-3)
a == b
```
--
```{r}
all(a == b)
```
--
```{r}
any(a == b)
```

---
# `all()` vs. `identical()`

--
```{r}
x <- c(1, 2, 3)
y <- c(1, 2, 3)
names(x) <- c('a', 'b', 'c')
names(y) <- c('one', 'two', 'three')
```
<br>

--
```{r}
all(x == y) # Only compares the elements
```
<br>

--
```{r}
identical(x, y) # Also compares the **names** of the elements
```
<br>

--
```{r}
names(y) <- c('a', 'b', 'c')
identical(x, y)
```


---
# Practice: Think-Pair-Share

[10 minutes](https://www.youtube.com/watch?v=4ASKMcdCc3g)

.pull-left[

Re-write `isPrime(n)` from last week, but **without loops!**

Remember, `isPrime(n)` takes a non-negative integer, `n`, and returns `TRUE` if it is a prime number and `FALSE` otherwise. Here's some test cases:

- `isPrime(1) == FALSE`
- `isPrime(2) == TRUE`
- `isPrime(7) == TRUE`
- `isPrime(13) == TRUE`
- `isPrime(14) == FALSE`
]
.pull-right[
Loop solution:

```{r, eval=FALSE}
isPrime <- function(n) {
    if (n <= 1) { return(FALSE) }
    if (n == 2) { return(TRUE) }
    for (i in seq(2, n-1)) {
        if (n %% i == 0) {
            return(FALSE)
        }
    }
    return(TRUE)
}
```
]

---
# 5 minute break - stand up, move around, stretch!

### [5 minutes](https://www.youtube.com/watch?v=_W0bSen8Qjg)

---
# Vectorized operations

--
```{r}
x1 <- c(1, 2, 3)
x2 <- c(4, 5, 6)
```
--
```{r}
# Addition
x1 + x2 # Returns (1+4, 2+5, 3+6)
```
--
```{r}
# Subtraction
x1 - x2 # Returns (1-4, 2-5, 3-6)
```
--
```{r}
# Multiplicattion
x1 * x2 # Returns (1*4, 2*5, 3*6)
```
--
```{r}
# Division
x1 / x2 # Returns (1/4, 2/5, 3/6)
```

---
# Vectorized operations

### Be careful to match dimensions!

--
```{r}
x1 <- c(1, 2, 3, 4)
x2 <- c(4, 5)
```
```{r, eval=FALSE}
x1 + x2
```
--
```{r, echo=FALSE}
x1 + x2
```

--
Single-value vector operations
```{r}
x1 <- c(1, 2, 3)
x2 <- c(4)
```
```{r, eval=FALSE}
x1 + x2
```
--
```{r, echo=FALSE}
x1 + x2
```

---
# Sorting
--
```{r}
a = c(2, 4, 6, 3, 1, 5)
a
```
--
```{r}
sort(a)
```
--
```{r}
sort(a, decreasing = TRUE)
```
--
Use `order()` to get the index values of the sorted order:
```{r}
order(a)
```

--
This does the same thing as `sort(a)`:
```{r}
a[order(a)]
```

---
# Group practice - no loops! [20 minutes](https://www.youtube.com/watch?v=kxGWsHYITAw)

1) `reverse(x)`

Write a function that returns the vector in reverse order. You cannot use the `rev()` function. Test cases:

- `reverse(c(5, 1, 3)) == c(3, 1, 5)`
- `reverse(c('a', 'b', 'c')) == c('c', 'b', 'a')`
- `reverse(c(FALSE, TRUE, TRUE)) == c(TRUE, TRUE, FALSE)`
- `reverse(seq(10)) == seq(10, 1, -1)`

2) `middleValue(a)`

Write a function that takes a vector of numbers `a` and returns the value of the middle element (or the average of the two middle elements). Test cases:

- `middleValue(c(0,0,0)) == 0`
- `middleValue(c(1,2,3)) == 2`
- `middleValue(c(4,5,6,7,8)) == 6`
- `middleValue(c(5,3,8,4)) == mean(c(3,8))`
- `middleValue(c(4,5,6,7)) == mean(c(5,6))`

---
# Group practice - no loops! [20 minutes](https://www.youtube.com/watch?v=kxGWsHYITAw)

1) `dotProduct(a, b)`

The "dot product" of two vectors is the sum of the products of the corresponding terms. So the dot product of the vectors `c(1,2,3)` and `c(4,5,6)` is `(1*4) + (2*5) + (3*6)`, or `4 + 10 + 18 = 32`. Write a function that takes two vectors and returns the dot product of those vectors. If the vectors are not equal length, ignore the extra elements in the longer vector. Test cases:

- `dotProduct(c(1,2,3), c(4,5,6)) == 32`
- `dotProduct(c(1,2), c(4,5,6)) == 14`
- `dotProduct(c(2,3,4), c(-7,1,9)) == 25`
- `dotProduct(c(0,0,0), c(-7,1,9)) == 0`

2) `alternatingSum(a)`

Write a function that takes a vector of numbers `a` and returns the alternating sum, where the sign alternates from positive to negative or vice versa. Test cases:

- `alternatingSum(c(5,3,8,4)) == (5 - 3 + 8 - 4)`
- `alternatingSum(c(1,2,3)) == (1 - 2 + 3)`
- `alternatingSum(c(0,0,0)) == 0`
- `alternatingSum(c(-7,5,3)) == (-7 - 5 + 3)`
