---
title: "Homework 4 - Strings"
output:
  html_document:
    number_sections: false
    toc: no
---

**Due**: Sunday, 20-Oct. at 8pm

**Rules**:

- Problems marked **SOLO** may not be worked on with other classmates, though you may consult instructors for help.
- For problems marked **COLLABORATIVE**, you may work in groups of up to 3 students who are in this course this semester (and then with nobody else except the course instructors). You may not split up the work -- everyone must work on every problem. And you may not simply copy any code but rather truly work together.
- Even though you work collaboratively, you still must submit your own solutions.

**Instructions**:

1) Before beginning this assignment, be sure to have read the [Strings](L7-strings.html) and [Vectors](L6-vectors.html) lessons.

2) Open RStudio and create a new project called "hw4-lastName", replacing "lastName" with your last name.

3) Download the [hw4.R template script](https://github.com/emse6574-gwu/2019-Fall/raw/gh-pages/hw_templates/hw4.R) and place it in RStudio project folder you just created.

4) Fill out your name, GW Net ID, and the names of anyone you worked with in the header of the "hw4.R" file.

5) Type all of your answers to the questions below in the "hw4.R" script.

6) After completing the questions, create a zip file of all files in your R project folder for this assignment.

7) Submit the zip file on Blackboard by the due deadline.

---

**Write solutions to the following functions in your "hw4.R" script.**

### 1) `reverseString(s)` [SOLO, 5 pts]

Write a function that returns the string in reverse order. So if `s` equals `"abcde"`,  `reverseString(s)` should equal `"edcba"`. You may assume that `s` only contains upper and/or lower case letters, but your solution must correctly return capital letters in their appropriate order. For example, `reverseString("aWordWithCaps")` should return `"spaChtiWdroWa"`.

### 2) `isPalindrome(s)` [SOLO, 5 pts]

Write a function that returns `TRUE` if the string `s` is a [Palindrome](https://en.wikipedia.org/wiki/Palindrome) and `FALSE` otherwise. The string `s` can contains any letter, number, or symbol, but it will be a character data type.

### 3) `sortString(s)` [SOLO, 10 pts]

Write the function `sortString(s)` that takes a string `s` and returns back an alphabetically sorted string. So `sortString("cba")` should return `"abc"`. You may assume that `s` only contains upper and/or lower case letters.

### 4) `vowelCount(s)` [SOLO, 10 pts]

Write the function `vowelCount(s)` that takes a string `s`, and returns the number of vowels in `s`, ignoring case (so `"A"` and `"a"` are both vowels). The vowels are `"a"`, `"e"`, `"i"`, `"o"`, and `"u"`. So, for example, `vowelCount("Abc def!!! a? yzyzyz!")` returns `3` (two `"a"`'s and one `"e"`).

### 5) `rotateStringLeft(s, k)` [COLLABORATIVE, 10 pts]

Write the function `rotateStringLeft(s, k)` that takes a string `s` and a non-negative integer `k`, and returns the string `s` rotated `k` places to the left. So, if `s = "iknowkungfu"` and `k = 2`, then the result should be `"nowkungfuik"`. If `k` is larger than the length of `s`, the function should continue to rotate the string beyond its starting point. So, if `s = "iknowkungfu"` and `k = 11`, then the result should be `"iknowkungfu"`, but if `k = 12`, the result should be `"knowkungfui"`.

### 6) `letterCount(s)` [COLLABORATIVE, 10 pts]

Write the function `letterCount(s)` that takes a string `s` and returns a named vector with the count of each letter in `s`, spanning the alphabet. If a letter does not appear in `s`, it should have a value of `0`. You may assume that `s` only contains upper and/or lower case letters. You should ignore cases, so `"a"` and `"A"` should be both treated as `"a"`.

For example, `letterCount("aaaaabbbc")` should return:

```{r, eval=FALSE}
a b c d e f g h i j k l m n o p q r s t u v w x y z
5 3 1 0 0 0 0 0 0 0 0 0 0 0 0 0 0 0 0 0 0 0 0 0 0 0
```

And `letterCount("someString")` should return:

```{r, eval=FALSE}
a b c d e f g h i j k l m n o p q r s t u v w x y z
0 0 0 0 1 0 1 0 1 0 0 0 1 1 2 0 0 1 2 1 0 0 0 0 0 0
```

### 7) `areAnagrams(s1, s2)` [COLLABORATIVE, 10 pts]

**Background**: An [anagram](https://en.wikipedia.org/wiki/Anagram) is a word or phrase formed by rearranging the letters of a different word or phrase, typically using all the original letters exactly once. With that in mind, write the function `areAnagrams(s1, s2)` that takes two strings, `s1` and `s2` (that you may assume contain only upper and/or lower case letters), and returns `TRUE` if the strings are anagrams, and `FALSE` otherwise. Treat `"a"` and `"A"` as the same letters (so `"Aba"` and `"BAA"` are anagrams). Hint: you can use either `sortString(s)` or `letterCount(s)` as a helper function - each are helpful for different approaches to solving this.

### 8) `isDigit(s)` [COLLABORATIVE, 10 pts]

Write the function `isDigit(s)` that takes a string `s` and returns `TRUE` if all characters in the string are numeric integers and there is at least one character, and `FALSE` otherwise. So `isDigit("123")` should return `TRUE`, `isDigit("123N")` should return `FALSE`, and `isDigit("")` should return `FALSE`.

### 9) `interleave(s1, s2)` [COLLABORATIVE, 15 pts]

Write the function `interleave(s1, s2)` that takes two strings, `s1` and `s2`, and interleaves their characters starting with the first character in `s1`. For example, `interleave('pto', 'yhn')` would return the string `"python"`. If one string is longer than the other, concatenate the rest of the remaining string onto the end of the new string. For example (`'a#'`, `'cD!f2'`) would return the string `"ac#D!f2"`. Assume that both `s1` and `s2` will always be strings.

### 10) `sameChars(s1, s2)` [COLLABORATIVE, 15 pts]

Write the function `sameChars(s1, s2)` that takes two strings and returns `TRUE` if the two strings are composed of the same characters (though perhaps in different numbers and in different orders); that is, if every character that is in the first string is in the second (and vice versa), and `FALSE` otherwise. This test is case-sensitive, so `"ABC"` and `"abc"` do not contain the same characters. The function returns `FALSE` if either parameter is not a string, but returns `TRUE` if both strings are empty.

---

### Bonus Credit 1) `hasBalancedParentheses(s)` [SOLO, 2.5 pts]

Write the function `hasBalancedParentheses(s)`, which takes a string `s` and returns `TRUE` if the parentheses in `s` are balanced and `FALSE` otherwise (ignoring all non-parentheses in the string). We say that parentheses are "balanced" if each right parenthesis closes (matches) an open (unmatched) left parenthesis, and no left parentheses are left unclosed (unmatched) at the end of the text. So, for example, `"( ( ( ) ( ) ) ( ) )"` is balanced, but `"( ) )"` is not balanced, and `"( ) ) ("` is also not balanced. Hint: keep track of how many right parentheses remain unmatched as you iterate over the string.

### Bonus Credit 2) `largestNumberInText(text)` [SOLO, 2.5 pts]

Write the function `largestNumberInText(text)` that takes a string of text and returns the largest integer value that occurs within that text, or `NULL` if no such value occurs. You may assume that the only numbers in the text are non-negative integers and that numbers are always composed of consecutive digits (without commas, for example). For example:

```{r, eval=FALSE}
largestNumberInText("I saw 3 dogs, 17 cats, and 14 cows!")
17 # The numeric value 17, not the string "17"
largestNumberInText("One person ate two hot dogs!")
NULL # Value NULL, not the string "NULL"
```

Hint: You may want to use `isDigit()` as a helper function!

---

**Page sources**:

Some content on this page has been modified from other courses, including:

- CMU [15-112: Fundamentals of Programming](http://www.kosbie.net/cmu/spring-17/15-112/), by [David Kosbie](http://www.kosbie.net/cmu/) & [Kelly Rivers](https://hcii.cmu.edu/people/kelly-rivers)
