---
title: "EMSE 6574: Programming for Analytics - Syllabus"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
library(knitr)
library(here)
library(kableExtra)

knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = '')
```

# Description

This course provides undergraduate students with a foundation in exploring data in the R programming language. Students will learn how to work with a wide variety of data types and structures as well as fundamental concepts in information visualization. The course will be managed on Github, and students will learn the basics of using git and Github to manage version controlling. All analyses will be conducted to support reproducibility from raw data to results using RMarkdown. There are no prerequisites, and no prior programming experience is necessary to succeed in this class.

# Learning objectives

Having successfully completed this course, students will be able to:

- Conduct a systematic exploratory data analysis (EDA) of different types of data.
- Understand fundamental principles of visualizing information for exploratory analysis and communication.
- Organize projects and manage version controlling with Git and GitHub
- [Wrangle data](https://r4ds.had.co.nz/wrangle-intro.html) from its original format into a fit-for-purpose format.
- Get data off the web and expose data, code, results on the web.
- Generate fully reproducible reports that contain contain code, equations, visualizations, and narrative text.

# Texts / References

All textbooks and references are available for free on the web, though physical copies can be purchased if desired:

- Data Wrangling in R:
    - Wickham, Hadley. "R for Data Science"
    [[free online](https://r4ds.had.co.nz/)],
    [[buy on amazon](https://www.amazon.com/Data-Science-Transform-Visualize-Model/dp/1491910399)]
- Data Visualization:
    - Healy, Kieran. "Data Visualization: A practical introduction"
    [[free online](http://socviz.co/)],
    [[buy on amazon](https://www.amazon.com/Data-Visualization-Introduction-Kieran-Healy/dp/0691181624)]
    - Wilke, Claus O. "Fundamentals of Data Visualization"
    [[free online](https://serialmentor.com/dataviz/)],
    [[buy on amazon](https://www.amazon.com/Fundamentals-Data-Visualization-Informative-Compelling/dp/1492031089)]
- Handy R [cheatsheets](https://www.rstudio.com/resources/cheatsheets/):
    - [RMarkdown](https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf)
    - [Data wrangling with the `dplyr` library](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf)
    - [Data visualization with the `ggplot2` library](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf)

# Preliminary Schedule

### Overview:
- PART I (Weeks 1-4): R, Git, and Data Management
- PART II (Weeks 5-9): Exploratory data analysis & data visualization
- PART III (Weeks 10-14): Reproducibility with RMarkdown

```{r echo=FALSE}
schedule <- read.csv(here::here('schedule.csv'), header=T)
kable(schedule) %>%
    kable_styling(bootstrap_options = c(
        'striped', 'hover', 'condensed', 'responsive'))
```

# Other resources

### Resources for finding datasets:
- [Kaggle](https://www.kaggle.com/datasets)
- [Google Dataset Search](https://toolbox.google.com/datasetsearch)
- [Tableau Community Forumss](https://community.tableau.com/docs/DOC-10635)

### Other texts:
- Evergreen, Stephanie,
  "Effective Data Visualization: The Right Chart for the Right Data"
  [[buy online](https://www.amazon.com/Effective-Data-Visualization-Right-Chart/dp/1544350880/)]

### Markdown:
- [Table generator](http://www.tablesgenerator.com/): Create tables and get export code for multiple formats.
- [Markdown It](https://markdown-it.github.io): Quickly demo Markdown code.
- [RMarkdown guide](http://www.stat.cmu.edu/~cshalizi/rmarkdown/)
- [Pimp my Rmd](https://holtzy.github.io/Pimp-my-rmd/)

### Plotting in `R`:
- [ggThemeAssist](https://github.com/calligross/ggthemeassist): Package for customizing plotting themes

### Tutorials:
- [Tidyverse in R](https://www.datacamp.com/community/tutorials/tidyverse-tutorial-r)
- [Transitioning from Excel to R](https://trendct.org/2015/06/12/r-for-beginners-how-to-transition-from-excel-to-r/)

### Inspiration:
- [An Incomplete List of Females in Data Visualization](https://stephanieevergreen.com/females-in-dataviz/), by Stephanie Evergreen
- [The Links Between Open Science and Star Wars](https://medium.com/read-write-participate/open-science-and-star-wars-2577b8081e8f)

### Examples of EDA in practice:
- [Death: Reality vs Reported](https://owenshen24.github.io/charting-death/), by Owen Shen
- [Is Usain Bolt really the fastest man on earth?](https://www.r-bloggers.com/fun-with-statistics-is-usain-bolt-really-the-fastest-man-on-earth/amp/), by Anindya Mozumdar
