---
title: "Course Syllabus"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
library(knitr)
library(here)
library(kableExtra)

knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = '')
```

# Course Information

| Name:         | EMSE 6574: Programming for Analytics, Section 10 |
|---------------|--------------------------------------------------|
| Semester:     | Fall 2019                                        |
| Meeting Time: | TBD                                              |
| Location:     | TBD                                              |

# Instructor Information

| Name:           | John Paul Helveston                     |
|-----------------|-----------------------------------------|
| Campus Address: | Science & Engineering Hall, Office 2830 |
| Campus Phone:   | +1 (202) 994-7173                       |
| Email:          | jph@gwu.edu                             |
| Office Hours:   | TBD                                     |

# Course Description

## GW Bulletin Description

Introduction to programming for data analytics using the [R computing language](https://www.r-project.org/); topics include R, RMarkdown, functions, conditionals, loops, debugging, testing, strings, file input/output, algorithmic thinking, coding style, efficiency, information visualization, Monte Carlo methods, webscraping data, packages, .

## Detailed Description

Introduction to programming for data analytics using the [R programming language](https://www.r-project.org/). Topics covered include introduction to the R programming language, structured program design in R, data structures for analytics in R, information visualization, file input/output, basic data analytics methods, data acquisition from real world sources. This course will prepare students for higher level courses in data analytics.

This course provides undergraduate students with a foundation in exploring data in the R programming language. Students will learn how to work with a wide variety of data types and structures as well as fundamental concepts in information visualization. The course will be managed on Github, and students will learn the basics of using git and Github to manage version controlling. All analyses will be conducted to support reproducibility from raw data to results using RMarkdown. There are no prerequisites, and no prior programming experience is necessary to succeed in this class.

## Prerequisites 

# Learning objectives

Having successfully completed this course, students will be able to:

- Conduct a systematic exploratory data analysis (EDA) of different types of data.
- Understand fundamental principles of visualizing information for exploratory analysis and communication.
- Organize projects and manage version controlling with Git and GitHub
- [Wrangle data](https://r4ds.had.co.nz/wrangle-intro.html) from its original format into a fit-for-purpose format.
- Get data off the web and expose data, code, results on the web.
- Generate fully reproducible reports that contain contain code, equations, visualizations, and narrative text.

# Texts / References

All textbooks and references are available for free on the web, though physical copies can be purchased if desired:

- Data Wrangling in R:
    - Wickham, Hadley. "R for Data Science"
    [[free online](https://r4ds.had.co.nz/)],
    [[buy on amazon](https://www.amazon.com/Data-Science-Transform-Visualize-Model/dp/1491910399)]
- Data Visualization:
    - Healy, Kieran. "Data Visualization: A practical introduction"
    [[free online](http://socviz.co/)],
    [[buy on amazon](https://www.amazon.com/Data-Visualization-Introduction-Kieran-Healy/dp/0691181624)]
    - Wilke, Claus O. "Fundamentals of Data Visualization"
    [[free online](https://serialmentor.com/dataviz/)],
    [[buy on amazon](https://www.amazon.com/Fundamentals-Data-Visualization-Informative-Compelling/dp/1492031089)]
- Handy R [cheatsheets](https://www.rstudio.com/resources/cheatsheets/):
    - [RMarkdown](https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf)
    - [Data wrangling with the `dplyr` library](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf)
    - [Data visualization with the `ggplot2` library](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf)

# Schedule

```{r echo=FALSE}
schedule <- read.csv(here::here('schedule.csv'), header=T)
kable(schedule) %>%
    kable_styling(bootstrap_options = c(
        'striped', 'hover', 'condensed', 'responsive', stripe_color = "#eeeeee"))
```