---
title: "Vectors"
output: html_document
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
rm(list=objects()) # start with a clean workspace
source("knitr_setup.R")
```

> ### Learning Objectives
>
> * Describe what a vector is.
> * Create vectors of different data types.
> * Use indexing to subset and modify specific portions of vectors.
> * Understand how to use vectorized functions to avoid loop operations.
>
> ### Suggested readings
>
> * [Chapter 20](https://r4ds.had.co.nz/vectors.html) of "R for Data Science", by Garrett Grolemund and Hadley Wickham
> * [Chapter 5.1](https://rstudio-education.github.io/hopr/r-objects.html#atomic-vectors) of "Hands-On Programming with R", by Garrett Grolemund

---

So far we've only dealt with objects that contain one value (e.g. `x <- 1`), but R actually stores those values in a _vector_ of length one:

```{r}
x <- 1
length(x)
is.vector(x)
```

A _vector_ is a basic data structure in R. All elements in a vector must have the same type.

> [Watch this 1-minute video for a quick summary of **vectors**](https://vimeo.com/220490316)

---

# Vector basics

## Creating vectors

The most basic way of creating a vector is to use the `c()` function ("c" is for "concatenate"):

```{r}
x <- c(1, 2, 3)
length(x)
```

You can also create a vector by using the `rep()` function, which replicates the same value `n` times:

```{r}
y <- rep(5, 10) # The number 5 ten times
z <- rep(10, 5) # The number 10 five times
```
```{r}
y
z
```

You can see how long a vector is using the `length()` function:

```{r}
length(y)
length(z)
```

Vectors can be of different types, but each element in the vector must have the _same_ type. If you mix types in a vector, R will coerce all the elements to either a numeric or character type:

```{r}
c(1, 2, "3") # Turns everything into a character
c(1, 2, TRUE, FALSE) # Turns everything into a number
c(TRUE, FALSE, "TRUE") # Turns everything into a character
```

## Deleting vectors

You can delete a vector by assigning `NULL` to it:

```{r}
x <- seq(1, 10)
x
x <- NULL
x
```

## Numeric vectors

As we saw in the [loops](5-loops.html) lesson, you can create a vector of integers using the `:` operator or the `seq()` function:

```{r}
x <- 1:10
y <- seq(1, 10)
```
```{r}
x
x == y
```

Numeric vectors don't all have to be integers though - they can be any numbers:

```{r}
v <- c(pi, 7, 42, 365)
v
typeof(v)
```

There are many functions that are designed to work on numeric vectors. Here are some common ones:

Function   |       Description           | Example
---------- | ----------------------------|------------------------------
`mean(x)` | Mean of values in `x` | `mean(c(1,2,3,4,5))` returns ``r mean(c(1,2,3,4,5))``
`median(x)` | Median of values in `x` | `median(c(1,2,2,4,5))` returns ``r median(c(1,2,2,4,5))``
`max(x)`  | Max element in `x` | `max(c(1,2,3,4,5))` returns ``r max(c(1,2,3,4,5))``
`min(x)`      | Min element in `x` | `min(c(1,2,3,4,5))` returns ``r min(c(1,2,3,4,5))``
`sum(x)`      | Sums the elements in `x` | `sum(c(1,2,3,4,5))` returns ``r sum(c(1,2,3,4,5))``
`prod(x)`     | Product of the elements in `x` | `prod(c(1,2,3,4,5))` returns ``r prod(c(1,2,3,4,5))``

## Character vectors

Character vectors are vectors where each element is a string:

```{r}
stringVector <- c('oh', 'what', 'a', 'beautiful', 'morning')
stringVector
typeof(stringVector)
```

As we'll see in the next lesson on [strings](L7-strings.html), you can "collapse" a character vector into a single string using the `str_c()` function from the `stringr` library:

```{r}
library(stringr)
str_c(stringVector, collapse = ' ')
```

## Logical vectors

Logical vectors contain only `TRUE` or `FALSE` elements:

```{r}
logicalVector <- c(rep(TRUE, 3), rep(FALSE, 3))
logicalVector
```

If you add a numeric type to a logical vector, the logical elements will be converted to either a `1` for `TRUE` or `0` for `FALSE`:

```{r}
c(logicalVector, 42)
```

**Warning**: If you add a character type to a logical vector, the logical elements will be converted to strings of `"TRUE"` and `"FALSE"`. So even though they may still _look_ like logical types, they aren't:

```{r}
y <- c(logicalVector, 'string')
y
typeof(y)
```

# Accessing elements in a vector

You can access elements from a vector using brackets `[]` and indices inside the brackets. You can use integer indices (probably the most common way), character indices (by naming each element), and logical indices.

## Using integer indices

Vector indices start from 1 (this is important - [most programming languages start from 0](https://en.wikipedia.org/wiki/Zero-based_numbering)):

```{r}
x <- seq(1, 10)
x[1] # Returns the first element
x[3] # Returns the third element
```

You can access multiple elements by using a vector of indices inside the brackets:

```{r}
x[c(1:3)]  # Returns the first three elements
x[c(2, 7)] # Returns the 2nd and 7th elements
```

You can also use negative integers to _remove_ elements, which returns all elements except that those specified:

```{r}
x[-1] # Returns everything except the first element
x[-c(2, 7)] # Returns everything except the 2nd and 7th elements
```

But you cannot mix positive and negative integers while indexing:

```{r error=TRUE}
x[c(-2, 7)]
```

If you try to use a float as an index, it gets rounded **down** to the nearest integer:

```{r error=TRUE}
x[3.1415] # Returns the 3rd element
x[3.9999] # Still returns the 3rd element
```

## Using characters indices

You can name the elements in a vector and then use those names to access elements. To create a named vector, use the `names()` function:

```{r error=TRUE}
names(x) <- c('one', 'two', 'three', 'four', 'five', 'six', 'seven', 'eight', 'nine', 'ten')
x
```

You can also create a named vector by putting the names directly in the `c()` function:

```{r error=TRUE}
x <- c('one' = 1, 'two' = 2, 'three' = 3)
x
```

Once your vector has names, you can then use those names as indices:

```{r error=TRUE}
x['one'] # Returns the first element
x[c('two', 'three')] # Returns the 2nd and 3rd elements
```

## Using logical indices

When using a logical vector for indexing, the position where the logical vector is `TRUE` is returned. This is helpful for filtering vectors based on conditions:

```{r}
x <- seq(1, 10)
x > 5 # Create logical vector
x[x > 5] # Put logical vector in brackets to filter out the TRUE elements
```

You can also use the `which()` operator to find the numeric indices for which a condition is `TRUE`, and then use those indices to select elements:

```{r}
which(x < 5) # Returns indices of TRUE elements
x[which(x < 5)] # Use which to select elements based on a condition
```

# Vectorized operations

Most base functions in R are "vectorized", meaning that when you give them a vector, they perform the operation on each element in the vector.

## Arithmetic operations

When you perform arithmetic operations on vectors, they are executed on an element-by-element basis:

```{r}
x1 <- c(1, 2, 3)
x2 <- c(4, 5, 6)
```
```{r}
# Addition
x1 + x2 # Returns (1+4, 2+5, 3+6)

# Subtraction
x1 - x2 # Returns (1-4, 2-5, 3-6)

# Multiplicattion
x1 * x2 # Returns (1*4, 2*5, 3*6)

# Addition
x1 / x2 # Returns (1/4, 2/5, 3/6)
```

When performing vectorized operations, the vectors need to have the same dimensions, or one of the vectors needs to be a single-value vector:

```{r error=TRUE}
# Careful! Mis-matched dimensions will only give you a warning, but will still return a value:
x1 <- c(1, 2, 3)
x2 <- c(4, 5)
x1 + x2
```

```{r}
# A single-value vector will still work:
x1 <- c(1, 2, 3)
x2 <- c(4)
x1 + x2
```

## Sorting

You can reorder the arrangement of elements in a vector by using the `sort()` function:

```{r}
a = c(2, 4, 6, 3, 1, 5)
sort(a)
sort(a, decreasing = TRUE)
```

To get the index values of the sorted order, use the `order()` function:

```{r}
order(a)
```

These indices tell us that the first value in the sorted arrangement of vector `a` is element number 5 (which is a `1`), the second value is element number `1` (which is a `2`), and so on. If you use `order()` as the indices to the vector, you'll get the sorted vector:

```{r}
a[order(a)] # Same as sort(a)
```

# Tips

**Use vector operations instead of a loop**:

As we saw in the [loops](5-loops.html) lesson, you can use a loop to perform an operation on each element in a vector. For example, the following loop prints out the decimal values for each element in a vector of floats:

```{r}
x <- c(3.1415, 1.618, 2.718)
for (i in 1:length(x)) {
    remainder <- x[i] %% 1
    print(remainder)
}
```

You could achieve the same thing by just performing the operation inside the loop (`x[i] %% 1`) on the whole vector:

```{r}
x %% 1
```

In many cases, using a vector can save you a whole lot of code by avoiding loops entirely!

---

**Page sources**:

Some content on this page has been modified from other courses, including:

- CMU [15-112: Fundamentals of Programming](http://www.kosbie.net/cmu/spring-17/15-112/), by [David Kosbie](http://www.kosbie.net/cmu/) & [Kelly Rivers](https://hcii.cmu.edu/people/kelly-rivers)
- Danielle Navarro's website ["R for Psychological Science"](https://psyr.org/index.html)
- [RStudio primers](https://rstudio.cloud/learn/primers/1.2)
