---
title: "Practice 2 - Programming Basics"
output:
  html_document:
    number_sections: false
    toc: no
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
rm(list=objects()) # start with a clean workspace
source("knitr_setup.R")
```

# Basic functions

Try out a few common, basic functions to learn what they do and how they work:

- Square root: `sqrt(64)` (you could also do this: `64^0.5`)
- Exponential: `exp(0)`, `exp(1)`, `exp(-10)`,
- Natural log: `log(10)`, log(1), `log(0)`, `log(-1)`, `log(exp(2))`
- Factorial: `factorial(1)`, `factorial(5)`, `factorial(-1)`, `factorial(0)`
- Round: `round(3.1415)`, `round(3.1415, digits=2)`
- Absolute value: `abs(-1)`, `abs(1)`

Here's a couple more to foreshadow where we'll be going:

- Type `rep("hello!", 100)` and take a look at the output. What does `rep()` do?
- Type `hist(x=rnorm(300))` and see what happens. What do you think the `hist()` function does? What about the `rnorm()` function? How could you find out more about each function?

# Objects

Make some objects using the following assignments:

```{r}
w <- 1
x <- 1
y <- 2
z <- 3
```

Using only these objects, try writing code to answer the following questions:

- Write a statement comparing the magnitude of `z` to `w` that returns `TRUE`
- Write a statement comparing the magnitude of `x` to `w` that returns `TRUE`
- Fill in the correct relational operators to make this statement return `TRUE`: `(x __ y) & (z __ y)`

# Data types

Try answering some of these questions **before** running any code - reason through the commands, then run the code and see if you were correct:

- What happens when you try to do math with a string: `"3" / 4`
- What do you think R will return if we divide infinity by a really big number: `Inf / 10^20`
- Why doesn't the following code return `TRUE` (Hint: [read ahead](L3-functions.html#7_tips) to find out):

```{r}
(0.1 + 0.2) == 0.3
```

---

**Page sources**:

Some content on this page has been modified from other courses, including:

- CMU [15-112: Fundamentals of Programming](http://www.kosbie.net/cmu/spring-17/15-112/), by [David Kosbie](http://www.kosbie.net/cmu/) & [Kelly Rivers](https://hcii.cmu.edu/people/kelly-rivers)
- Danielle Navarro's website ["R for Psychological Science"](https://psyr.org/index.html)
