---
title: "Practice 3 - Functions"
output:
  html_document:
    number_sections: false
    toc: no
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
rm(list=objects()) # start with a clean workspace
source("knitr_setup.R")
```

**Hint**: Read the [Getting Started Tips](L1.2-getting-started.html#tips), [Programming Basics Tips](L1.3-programming-basics.html#tips), and [Functions Tips](L3.1-functions.html#tips) - these could come in handy!

# Mystery Functions

In just a few words of plain English, what does each of the following functions do? Provide some explanation for your answer.

```{r, eval=FALSE}
f <- function(x, y) {
    epsilon = 0.0000001
    d1 = abs(x)^2
    d2 = abs(y)^0.5
    return(abs(d1 - d2) < epsilon)
}

g <- function(x) {
    return(is.integer(x) & (x %/% 10 == x %% 10))
}

h <- function(x) {
    return(is.integer(x) & (x >= 0) & ((as.integer(round(x^0.5)))^2 == x))
}
```

# Write functions

Here are a bunch of functions you should be able to write. Any of these may appear (directly or modified) on a quiz or exam!

### `celsiusToFahrenheit(celsius)`

Given a temperature in Celsius, return the same temperature in Fahrenheit.  So `celsiusToFahrenheit(0)` returns `32`, and `celsiusToFahrenheit(100)` returns `212`.

### `isPerfectCube(x)`

Given an integer value `x`, returns `TRUE` if it is a perfect cube and `FALSE` otherwise. That is, return `TRUE` if there is another integer `y` such that `x = y^3`. Thus, `isPerfectCube(27)` returns `TRUE`, but `isPerfectCube(16)` returns `FALSE`.

### `kthDigit(x, k)`

Given two integers, `x` and `k`, return the kth digit of `x`, counting from the right. So:

- `kthDigit(789, 0)` returns `9`
- `kthDigit(789, 1)` returns `8`
- `kthDigit(789, 2)` returns `7`
- `kthDigit(789, 3)` returns `0`

Negative numbers should work, too, so `kthDigit(-789, 0)` returns `9`.

### `nthFibonacciNumber(n)`

Write the function `nthFibonacciNumber(n)` that takes a positive integer `n` and (without using loops) returns the nth [Fibonacci number](https://en.wikipedia.org/wiki/Fibonacci_number){target="_blank"}. So:

- `nthFibonacciNumber(1)` returns `1`
- `nthFibonacciNumber(2)` returns `1`
- `nthFibonacciNumber(3)` returns `2`
- `nthFibonacciNumber(4)` returns `3`
- `nthFibonacciNumber(5)` returns `5`
- `nthFibonacciNumber(6)` returns `8`

### `isTriangularNumber(n)`

A number is triangular if it equals `1+2+3+...+N` for some integer `N`. Given an integer value `n`, return `TRUE` if it is triangular, and `FALSE` otherwise. Note that this relates to the pool ball problems. Hint: For this problem you should research [Triangular Numbers](https://en.wikipedia.org/wiki/Triangular_number){target="_blank"}. Hint 2: This is directly related to the `numberOfPoolBalls(rows)` homework problem!

### `dcHour(londonHour)`

This function takes an integer, the current hour in London, and returns the current hour in Washington DC (which is 5 hours behind London). However, London time is given in 24-hour time (so `londonHour` is between 0 and 23, inclusive), but Washington DC time must be returned in 12-hour time (so the result must be between 1 and 12, inclusive, where "am" and "pm" are ignored). Here are some examples:

<div class="tableSmall">
`londonHour` | `dcHour(londonHour)`
-------------| -------------------
0 (midnight) | 7 (7pm)
10 (10am)    | 5 (5am)
12 (noon)    | 7 (7am)
17 (5pm)     | 12 (noon)
18 (6pm)     | 1 (1pm)
</div>

Note: the hardest part is when it is 12 o'clock in DC.

---

**Page sources**:
Some content on this page has been modified from other courses, including:

- CMU [15-112: Fundamentals of Programming](http://www.kosbie.net/cmu/spring-13/15-112/)
