---
title: "Practice 4 - Conditionals & Testing"
output: html_document
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
rm(list=objects()) # start with a clean workspace
source("knitr_setup.R")
```

# Code tracing

### 1)
After running this code, what is the value of `a` and `b`?

```{r, eval = FALSE}
f <- function(x) {
    x = x + 1
    if ((x %% 2) == 0) {
        x = x - 1
    }
    x = 2*x
}

a <- f(7)
b <- f(12)
```

### 2)
Write the output of this code by hand:

```{r, eval = FALSE}
f <- function(x) {
    if ((x %% 3) == 0) {
        cat('woo!')
        return(x %/% 3)
    }
    return(x %% 2)
}

cat(f(9))
cat(f(11))
```

### 3)
Write the output of this code by hand:

```{r, eval = FALSE}
f <- function(x) {
    if (x > 0) {
        cat('abc')
        x = 2*x
    } else if (x <= 0) {
        x = abs(x)
        cat('cba')
    }
    cat(x)
}

cat(f(-9))
cat(f(15))
```

# Write functions

Here are some functions you should be able to write. Any of these may appear (directly or modified) on a quiz or exam! Your function should be able to pass the test functions provided.

## `isPositiveMultipleOf4Or7(n)`

### a)

Write the function `isPositiveMultipleOf4Or7(n)` that returns `TRUE` if `n` is a positive multiple of 4 or 7 and `FALSE` otherwise. Note than `n` could be any data type.

### b)

Write the test function `testIsPositiveMultipleOf4Or7()` that tests the function `isPositiveMultipleOf4Or7(n)` for a variety of values of `n`. Consider cases that you might not expect, such as cases where `n` is not a number.
