---
title: "Resources"
output: html_document
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
rm(list=objects()) # start with a clean workspace
source("knitr_setup.R")
```

---

# Getting Help

```{r child = here::here('syllabus', 'getting-help.Rmd')}
```

---

# Doing stuff in `R`

## Data Wrangling

- Wickham, Hadley. "R for Data Science"
[[free online](https://r4ds.had.co.nz/){target="_blank"}],
[[buy on amazon](https://www.amazon.com/Data-Science-Transform-Visualize-Model/dp/1491910399){target="_blank"}]

## Data Visualization

- Healy, Kieran. "Data Visualization: A practical introduction"
[[free online](http://socviz.co/){target="_blank"}],
[[buy on amazon](https://www.amazon.com/Data-Visualization-Introduction-Kieran-Healy/dp/0691181624){target="_blank"}]
- Wilke, Claus O. "Fundamentals of Data Visualization"
[[free online](https://serialmentor.com/dataviz/){target="_blank"}],
[[buy on amazon](https://www.amazon.com/Fundamentals-Data-Visualization-Informative-Compelling/dp/1492031089){target="_blank"}]
- [ggThemeAssist](https://github.com/calligross/ggthemeassist): Package for customizing plotting themes

## RStudio [cheatsheets](https://www.rstudio.com/resources/cheatsheets/)

- [RMarkdown](https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf){target="_blank"}
- [Data wrangling with the `dplyr` library](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf){target="_blank"}
- [Data visualization with the `ggplot2` library](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf){target="_blank"}

## RMarkdown

- [60 second markdown guide](https://commonmark.org/help/){target="_blank"}
- [10 minute markdown tutorial](https://commonmark.org/help/tutorial/){target="_blank"}
- [Markdown It](https://markdown-it.github.io){target="_blank"}: Quickly demo Markdown code.
- [Table generator](http://www.tablesgenerator.com/){target="_blank"}: Create tables and get export code for multiple formats.
- [CMU RMarkdown guide (detailed)](http://www.stat.cmu.edu/~cshalizi/rmarkdown/){target="_blank"}

---

# Datasets

- [Kaggle](https://www.kaggle.com/datasets){target="_blank"}
- [Google Dataset Search](https://toolbox.google.com/datasetsearch){target="_blank"}
- [Tableau Community Forumss](https://community.tableau.com/docs/DOC-10635){target="_blank"}
- [More datasets for teaching data science: The expanded `dslabs` package](https://simplystatistics.org/2019/07/19/more-datasets-for-teaching-data-science-the-expanded-dslabs-package/){target="_blank"}, by Amy Gill

---

# Helpful Tutorials

- [RStudio Primers](https://rstudio.cloud/learn/primers){target="_blank"}
- [Tidyverse in R](https://www.datacamp.com/community/tutorials/tidyverse-tutorial-r){target="_blank"}
- [Transitioning from Excel to R](https://trendct.org/2015/06/12/r-for-beginners-how-to-transition-from-excel-to-r/){target="_blank"}

---

# Inspiration

- [The Links Between Open Science and Star Wars](https://medium.com/read-write-participate/open-science-and-star-wars-2577b8081e8f){target="_blank"}
