---
title: "EMSE Undergraduate Data Analytics"
author: "John Paul Helveston"
date: December 09, 2019
output:
  xaringan::moon_reader:
    css: ['default', 'https://emse6574-gwu.github.io/2019-Fall/classNotes/classNotes.css']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "https://platform.twitter.com/widgets.js"
    seal: false
---

```{r setup, include=FALSE}
library(knitr)
library(here)
library(tidyverse)
library(cowplot)
library(ggrepel)
library(fontawesome) # from github: https://github.com/rstudio/fontawesome
options(
    htmltools.dir.version = FALSE,
    knitr.table.format = "html",
    knitr.kable.NA = '')
knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    fig.retina = 3)
set.seed(42)
```

name: xaringan-title
class: title-slide, center, middle, inverse

# EMSE Undergraduate Data Analytics

.large[John Paul Helveston | December 09, 2019]

---
class: center 

# Translating _data_ into _information_

--
.leftcol[
**Data**

Student engagement scores

 Class       | School Type | Male | Female 
 ------------|-------------|------|-------
 Special Ed. | Charter     | 643  | 793
 Special Ed. | Public      | 735  | 928
 General Ed. | Charter     | 590  | 724
 General Ed. | Public      | 863  | 662   

]
--
.rightcol[
**Information**

.blackborder[
```{r, echo=FALSE, fig.width=6, fig.height=5}
engagement <- data.frame(
    Male   = c(643, 735, 590, 863),
    Female = c(793, 928, 724, 662),
    School = c('Special Ed., Charter', 'Special Ed., Public',
               'General Ed., Charter', 'General Ed., Public'),
    Highlight = c(0, 0, 0, 1)) %>%
    gather(Gender, Engagement, Male:Female) %>%
    mutate(Gender = fct_relevel(Gender, c('Male', 'Female')))
engagement$x <- rep(c(0, 1), each = 4)
engagement_plot <- engagement %>%
    mutate(Highlight = as.factor(Highlight)) %>% 
    ggplot(aes(x=x, y=Engagement, group=School, color=Highlight)) +
    geom_point() +
    geom_line() +
    scale_color_manual(values = c('#757575', '#ed573e')) +
    labs(x='Sex', y='Engagement') +
    theme_cowplot() +
    scale_x_continuous(limits = c(-0.2, 2), labels = c('Male', 'Female'),
                       breaks = c(0, 1)) +
    theme(legend.position = 'none') +
    geom_text_repel(aes(label = School, color = as.factor(Highlight)),
                    data          = subset(engagement, Gender == 'Female'),
                    size          = 5,
                    nudge_x       = 0.2,
                    hjust         = 0,
                    segment.color = NA)
engagement_plot
```
]]

---
# Data exploration: an iterative process

--
.leftcol[
Encode data:

.code60[
```{r}
engagement_data <- data.frame(
    Male   = c(643, 735, 590, 863),
    Female = c(793, 928, 724, 662),
    School = c('Special Ed., Charter', 'Special Ed., Public',
               'General Ed., Charter', 'General Ed., Public'))
engagement_data
```
]]
--
.rightcol[
Re-shape data for plotting:

.code60[
```{r}
engagement_data <- engagement_data %>% 
    gather(Gender, Engagement, Male:Female) %>%
    mutate(Gender = fct_relevel(
      Gender, c('Male', 'Female')))
engagement_data
```
]]

---
# Data exploration: an iterative process

.leftcol[
Initial exploratory plotting:

.code60[.blackborder[
```{r, fig.width=6, fig.height=3}
engagement_data %>% 
    ggplot() +
    geom_bar(aes(x = School, y = Engagement, 
                 fill = Gender), stat = 'identity', 
             position = 'dodge') +
    coord_flip()
```
]]]
--
.rightcol[
More exploratory plotting:

.blackborder[
```{r, echo=FALSE, fig.width=6, fig.height=5}
engagement %>%
    mutate(Highlight = as.factor(Highlight)) %>% 
    ggplot(aes(x=x, y=Engagement, group=School, color=School)) +
    geom_point() +
    geom_line() +
    labs(x='Sex', y='Engagement') 
```
]]

---
# Data exploration: an iterative process

.leftcol[
Directly label figure:

.blackborder[
```{r, echo=FALSE, fig.width=6, fig.height=5}
engagement %>%
    mutate(Highlight = as.factor(Highlight)) %>% 
    ggplot(aes(x=x, y=Engagement, group=School, color=School)) +
    geom_point() +
    geom_line() +
    labs(x='Sex', y='Engagement') +
    theme_cowplot() +
    scale_x_continuous(limits = c(-0.2, 2), labels = c('Male', 'Female'),
                       breaks = c(0, 1)) +
    theme(legend.position = 'none') +
    geom_text_repel(aes(label = School, color = as.factor(School)),
                    data          = subset(engagement, Gender == 'Female'),
                    size          = 5,
                    nudge_x       = 0.2,
                    hjust         = 0,
                    segment.color = NA)
```
]]
--
.rightcol[

Format colors:

.blackborder[
```{r, echo=FALSE, fig.width=6, fig.height=5}
engagement_plot 
```
]]

---
# A fully reproducible analysis

.leftcol[
Code:

.code50[
```{r, eval=FALSE}
engagement_data <- data.frame(
    Male   = c(643, 735, 590, 863),
    Female = c(793, 928, 724, 662),
    School = c('Special Ed., Charter', 'Special Ed., Public',
               'General Ed., Charter', 'General Ed., Public'),
    Highlight = c(0, 0, 0, 1)) %>%
    gather(Gender, Engagement, Male:Female) %>%
    mutate(Gender = fct_relevel(Gender, c('Male', 'Female')))
engagement_data %>%
    mutate(Highlight = as.factor(Highlight)) %>% 
    ggplot(aes(x=x, y=Engagement, group=School, color=Highlight)) +
    geom_point() +
    geom_line() +
    scale_color_manual(values = c('#757575', '#ed573e')) +
    labs(x='Sex', y='Engagement') +
    theme_cowplot() +
    scale_x_continuous(limits = c(-0.2, 2), labels = c('Male', 'Female'),
                       breaks = c(0, 1)) +
    theme(legend.position = 'none') +
    geom_text_repel(aes(label = School, color = as.factor(Highlight)),
                    data          = subset(engagement, Gender == 'Female'),
                    size          = 5,
                    nudge_x       = 0.2,
                    hjust         = 0,
                    segment.color = NA)
```
]]
.rightcol[
Result:

.blackborder[
```{r, echo=FALSE, fig.width=6, fig.height=5}
engagement_plot 
```
]]

---
## Course 1: [Programming for Analytics](https://emse6574-gwu.github.io/2019-Fall/)

**"Computational Literacy"**
- Data types (numerical, logical, strings), conditional statements (if/else), loops, functions, testing.
- Data structures, data import / export, data manipulation & visualization.
<br>

--
<br>

## Course 2: [Exploratory Data Analysis](https://emse-eda-gwu.github.io/2020-Spring/)

**"Data Literacy"**
- Data provenance, validity.
- Design rules for visualizing and communicating information contained in data.
- Reproducibility: Reports that contain code, equations, visualizations, and narrative text.